Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-06-29 agujournaltemplate.tex
Reviewed Method and Results

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
\caption{Bathimetric map of the CARIACO Basin in the Caribbean Ocean, off the coast of Venezuela, indicating the CARIACO Mooring, the station of the Ocean Time-Series (located at \ang{10.50}N, \ang{64.67}W). A sill, running almost parallel to the coast at an average depth of 100\,m, from the Isla de Magarita and the Isla de la Tortuga and beyond, encloses the Cariaco basin from the North.}
\label{fig:map}
\end{figure}

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time Series Program by means of monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this work. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela. Details about the sampling procedures adopted in CARIACO, sampling schemes, analytical methods,  data quality control and inter-calibration procedures, have been already described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to  present by Kent Fanning (University of South Florida, St. Petersburg, USA). Samples from eight cruises were analyzed by both laboratories. Data for PO$_4$ and SiO$_4$ were in strong agreement (within ±5\%), whereas  NO$_\text{x}$ and NH$_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. We used a merged dataset for all nutrients, in which coincidental monthly measurements are averaged.

@@

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}. We extracted the -u wind component at 10\,m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate for total precipitation, comprising all water that falls on Earth's surface as rain or snow.

@@

    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to create a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}. Furthermore, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then it averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated over a depth of 100 m for observed genus counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO]}_3$, $\mathrm{[PO]}_4$ and $\mathrm{[SiO]}_4$). 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. 
    We show a model run without any time lags in  supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix). 
    % Note: Perhaps add some more info about Supplmentary Results and test runs here?

@@

% Figure 4
We undertook clustering analysis on annually aggregated cell counts from microscopy and on individual monthly measurements. Figure \ref{fig:clustering}$a$ shows the results of clustering the annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix. The distinct clusters clearly emerge from these analysis, , the first cluster encompassing the years 1996-2004 and 2014-2016 and the second cluster including the intermediate years from 2004 to 2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, it reveals an increasing trend in turnover from 1996 to 2004. Turnover subsequently decreases during the period underpinning cluster 2 and increases again in 2014, marking a return to cluster 1. This pattern is also reflected in the hierarchical clustering results, Figure \ref{fig:clustering}$a$, where cluster 2 shows a lower distance in the clustering tree between the years. We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster, respectively. The diatom \textit{Thalassiosira} dominated the early period and was abundant in all other periods. \textit{Pseudo-nitzschia} dominates cluster 2, with the haptophyte \textit{Emiliana} being the second most abundant genus. The later part of cluster 1 is dominated by \textit{Emiliana}.

To check whether the annual aggregation of community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (Figure \ref{fig:clustering}$c$). The NMDS clustering is overlaid with environmental data and climate indices in the attempt of unveiling some of the underlying trends. The horizontal axis separates the two clusters, with cluster 2 located over positive anomalies in AMO and precipitation and cluster 1 located over positive anomalies in evaporation, salinity, and MEI v.2. The second NMDS axis appears to follow the trends in upwelling intensity, as indicated by the arrows poinintg in opposite directions for nutrient concentrations, temperature, and wind speed. Overall, the NMDS clustering confirms the pattern shown by clustering on annually aggregated data.

@@

% Figure 5 - NEW
Figure \ref{fig:clustcomp} shows density distributions of environmental variables, chlorophyll \textit{a}, and diversity indices for the two clusters. Upwelling-related variables show distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} isotherm, lower temperatures, reduced evaporation, and higher salinity and nutrient concentrations, particularly for nitrate. The AMO index shows a pronounced separation between the two clusters, with cluster 1 exhibiting more negative anomalies, while cluster 2 showing higher occurrence of positive anomalies. Although there are no significant differences in genus richness and the Shannon index between clusters, there is a slightly increased median in the Pielou index for cluster 2. The density distribution reveals some differences. Cluster 1 shows higher variance for genus richness, while cluster 2  shows greater variance and a bimodal distribution for Shannon and Pielou indices.

@@

% Figure 6
To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were used to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. Figure \ref{fig:GF}$a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity. The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentrations scored the lowest in predicting changes within the community. Gradient Forest predicts shifts in the community by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF}$b$. The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are from diatoms. However, two haptophyte and three dinoflagellate genera rank relatively high.
Figure \ref{fig:GF}$c$ shows the cumulative importance of the predictor variables across the parameter range. The black, dashed lines indicate cumulative importance across the entire community, and the solid, dashed lines indicate importance per phytoplankton group. For the AMO index, we see a clear pattern in which negative anomalies affect the community. This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria. Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition. Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows esponse curves for all variables not shown here. For wind speed, SST, and salinity, changes occurred mostly within parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}), and high salinity (above \qty{36.8}{PSU}).

@@

\begin{table}
\caption{Results of two-sided Wilcoxon sum rank test with continuity correction for variables between the two clusters. Cluster 1 covers the periods 1996-2003 and 2014-2016 and cluster 2 covers the period 2004-2013.}
\centering
\begin{tabular}[t]{lrrl}
\toprule
Variable & W & p.value & Significance\\
\midrule
AMO & 3494.0 & 1.86e-14 & ***\\
MEI v.2 & 9523.5 & 0.006 & **\\
Wind speed & 7074.0 & 0.143 & n.s.\\
Precipitation & 6251.0 & 0.004 & **\\
Evaporation & 9979.0 & 3.68e-5 & ***\\
\addlinespace
21 °C Isotherm & 5293.0 & 0.034 & *\\
SST & 4712.0 & 0.001 & ***\\
Salinity & 7636.0 & 0.002 & **\\
$NO_3$ & 6378.0 & 0.018 & *\\
$PO_4$ & 7359.0 & 0.012 & *\\
$SiO_4$ & 5998.0 & 0.736 & n.s.\\
\addlinespace
Chlorophyll $a$ & 8229.0 & 9.63e-6 & ***\\
Genus Richness & 5913.0 & 0.954 & n.s.\\
Shannon Index & 5142.0 & 0.108 & n.s.\\
Pielou Index & 4874.0 & 0.029 & *\\
\bottomrule
\end{tabular}
\end{table}

>>>>

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
\caption{Bathimetric map of the CARIACO Basin in the Caribbean Ocean, off the coast of Venezuela, indicating the CARIACO Mooring, the station of the Ocean Time-Series (located at \ang{10.50}N, \ang{64.67}W). A sill, running almost parallel to the coast at an average depth of \qty{100}{m}, from the Isla de Magarita and the Isla la Tortuga and beyond, encloses the Cariaco basin from the North.}
\label{fig:map}
\end{figure}

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time Series Program by means of monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Isla de Magarita, Venezuela. Details about the sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to  present by Dr. K. Fanning (University of South Florida, St. Petersburg, USA). Samples from eight cruises were analyzed by both laboratories. Data for PO$_4$ and SiO$_4$ were in strong agreement (within ±5\%), whereas  NO$_\text{x}$ and NH$_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. We used a merged dataset for all nutrients, in which coincidental monthly measurements are averaged.

@@

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}. We extracted the -u wind component at \qty{10}{\meter} from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate for total precipitation, comprising all water that falls on Earth's surface as rain or snow.

@@

    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to create a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}. Furthermore, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then it averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated over a depth of 100 m for observed genus counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO]}_3$, $\mathrm{[PO]}_4$ and $\mathrm{[SiO]}_4$). 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. 
    We show a model run without any time lags in  supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix).

@@

% Figure 4
We undertook clustering analysis on annually aggregated cell counts from microscopy and on individual monthly measurements. Figure \ref{fig:clustering}$a$ shows the results of clustering the annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix. Two distinct clusters emerge from this analysis, with the first cluster encompassing the years 1996-2004 and 2014-2016, and the second cluster including the intermediate years from 2005 to 2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, it reveals an increasing trend in turnover from 1996 to 2004. Turnover subsequently decreases during the period underpinning cluster 2 and increases again in 2014, marking a return to cluster 1. This pattern is also reflected in the hierarchical clustering results, Figure \ref{fig:clustering}$a$, where cluster 2 shows a lower distance in the clustering tree between the years. We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster, respectively. The diatom \textit{Thalassiosira} dominated the early period and was abundant in all other periods. \textit{Pseudo-nitzschia} dominates cluster 2, with the haptophyte \textit{Emiliana} being the second most abundant genus. The later part of cluster 1 is dominated by \textit{Emiliana}.

To check whether the annual aggregation of community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (Figure \ref{fig:clustering}$c$). The NMDS clustering is overlaid with environmental data and climate indices in the attempt of unveiling some of the underlying trends. The horizontal axis separates the two clusters, with cluster 2 located over positive anomalies in AMO and precipitation and cluster 1 located over positive anomalies in evaporation, salinity, and MEI v.2. The second NMDS axis appears to follow the trends in upwelling intensity, as indicated by the vectors pointing in opposite directions for nutrient concentrations, temperature, and wind speed. Overall, the NMDS clustering confirms the pattern shown by clustering on annually aggregated data.

@@

% Figure 5 - NEW
Figure \ref{fig:clustcomp} shows density distributions of environmental variables, chlorophyll \textit{a}, and diversity indices for the two clusters. Upwelling-related variables exhibit distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} isotherm, lower temperatures, reduced evaporation, and higher salinity and nutrient concentrations, particularly for nitrate. The AMO index shows a pronounced separation between the two clusters, with cluster 1 exhibiting more negative anomalies, while cluster 2 shows a higher occurrence of positive anomalies. Although there are no significant differences in genus richness and the Shannon index between clusters, there is a slightly increased median in the Pielou index for cluster 2 (Table \ref{tab:ClustCompWilcox}). The density distribution reveals some differences. Cluster 1 shows higher variance for genus richness, while cluster 2 shows greater variance and a bimodal distribution for the Shannon and Pielou indices.

@@

% Figure 6
To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were used to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. Figure \ref{fig:GF}$a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity. The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentrations scored the lowest in predicting changes within the community. Gradient Forest predicts shifts in the community by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF}$b$. The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are from diatoms. However, two haptophyte and three dinoflagellate genera rank relatively high.
Figure \ref{fig:GF}$c$ shows the cumulative importance of the predictor variables across the parameter range. The black, dashed lines indicate cumulative importance across the entire community, and the solid, dashed lines indicate importance per phytoplankton group. For the AMO index, we see a clear pattern in which negative anomalies affect the community. This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria. Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition. Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows response curves for all variables not shown here. For wind speed, SST, and salinity, changes occurred mostly within parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}), and high salinity (above \qty{36.8}{PSU}).

@@

\begin{table}
\caption{Results of two-sided Wilcoxon sum rank test with continuity correction for variables between the two clusters. Cluster 1 covers the periods 1996-2004 and 2014-2016 and cluster 2 covers the period 2005-2013.}
\centering
\begin{tabular}[t]{lrrl}
\toprule
Variable & W & p.value & Significance\\
\midrule
AMO & 3494.0 & 1.86e-14 & ***\\
MEI v.2 & 9523.5 & 0.006 & **\\
Wind speed & 7074.0 & 0.143 & n.s.\\
Precipitation & 6251.0 & 0.004 & **\\
Evaporation & 9979.0 & 3.68e-5 & ***\\
\addlinespace
21 °C Isotherm & 5293.0 & 0.034 & *\\
SST & 4712.0 & 0.001 & ***\\
Salinity & 7636.0 & 0.002 & **\\
$NO_3$ & 6378.0 & 0.018 & *\\
$PO_4$ & 7359.0 & 0.012 & *\\
$SiO_4$ & 5998.0 & 0.736 & n.s.\\
\addlinespace
Chlorophyll $a$ & 8229.0 & 9.63e-6 & ***\\
Genus Richness & 5913.0 & 0.954 & n.s.\\
Shannon Index & 5142.0 & 0.108 & n.s.\\
Pielou Index & 4874.0 & 0.029 & *\\
\bottomrule
\end{tabular}

\label{tab:ClustCompWilcox}
\end{table}
